Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-11-14 revision/review-response.Rmd

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
csl: ../florida-scientist.csl
bibliography: ../refs.bib
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(message = F, echo = F, cache = F, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files(here::here('revision/review-response.Rmd'))
```

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below. In general, most of the revisions address concerns...

All page and line numbers in the reviewer comments refer to the original submission.  Page numbers noted in our response refer to the revised submission.

## Reviewer 1

1) Does this paper contain new knowledge, or new interpretation of existing knowledge?

The authors developed a rapid-response dashboard for the purposes of monitoring the effects of the Piney Point wastewater release in Summer of 2021. The collaborative, multi-organizational approach to the response, data collection, and subsequent dissemination of important information concerning the release and its effects on the health of Tampa Bay is certainly a novel and important product.

2)  Does it duplicate in any substantial way material published elsewhere?

The paper does not duplicate material published. While data from the Piney Point response have been analyzed and published in other journals, this paper serves as a detailed roadmap for how the data were collected and wrangled into a useable, public-facing product.

3)  Are technical terms used and spelled correctly?

Yes, all technical terms are used and spelled correctly.

4) Are there major flaws in the study design, execution or analysis?

There are no major flaws in the methods used to produce the Piney Point response dashboard.

5)  Please list below any suggestions for improvement.

A brief review for grammar/spelling/punctuation may be warranted. For example, “e.g.” statements should never end in “etc” as the e.g. is intended for examples (e.g., page 6). I imagine that a copy-edit review prior to publication will catch most of these. However, the paper overall was well-written and structured in a logical, easy-to-follow manner.

Table 2 may need to be reformatted to adjust spacing (hanging ‘a’ in Pennsylvania).

* __Response__: Thank you for your comments and we are pleased you found the manuscript a valuable contribution.  We have proofed the revision for any grammar/spelling/punctuation issues, specifically addressing your comment about "e.g." statements.  The issue noted for Table 2 will be addressed during copy-editing as this is the unformatted version that will be changed on publication.

## Reviewer 2

<!-- # References -->

### 2022-11-14 manu-draft.Rmd
starting revision

The long-term recovery of Tampa Bay, Florida represents a collaborative, multi-stakeholder effort to quantify and limit external nitrogen loading as a key pollutant impacting ecosystem health.  Nitrogen loading to the estuary has been reduced by approximately two thirds of historical conditions primarily through upgrades to wastewater treatment systems, stormwater controls, and habitat restoration projects [@Johansson91;@Lewis98;@Greening14;@Beck19].  As a result, chlorophyll-a concentrations (a proxy for phytoplankton production) have decreased, water clarity has improved, and seagrasses recovered to a historical 1950s baseline in 2016 [@Greening14;@Sherwood17].  Despite these historical gains, ongoing management challenges for protecting and restoring the natural resources of Tampa Bay persist (e.g. continuing stressors associated with climate change and increasing watershed development).  Recent seagrass losses over the 2018 to 2022 period and seasonally recurring harmful algal blooms since the mid-2000s in portions of Tampa Bay highlight contemporary challenges affecting the bay's current ecology [@Lopez16;@Tomasko20].

@@

Water quality monitoring focused on parameters indicative of elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily on total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton production, and additional parameters expected to respond to nutrient loading (e.g., secchi depth, dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected at transect locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay. Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet cover-abundance estimates at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

@@

A second synthesis challenge was unifying the multiple data types hosted on Google Drive into a usable format for the dashboard and additional analyses (Figure \@ref(fig:synthesis), data synthesis).  Several folders on Google Drive were created that included relevant files for each type of monitoring data (e.g., water quality, phytoplankton, etc.).  Each folder typically contained dozens of files saved as Google Sheets, often uploaded as a single sampling event by an individual monitoring agency.  For example, the folder for the water quality sampling included 91 separate files.  Extensive data "wrangling" was required to combine the files into a tidy format [@Wickham14c] and a custom processing script was created by TBEP using the open source R statistical programming language [@RCT20]. An RStudio project was created with full version control that included the synthesis script, `dat_proc.R`, that imported and processed all files from Google Drive into individual binary \.RData files (i.e., individual file for water quality, phytoplankton, etc.). These discrete files were later used to develop a subsequent results dashboard .  Several R packages were indispensable in the data synthesis effort and these included the 1) googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages for connecting and downloading files from Google Drive, and 2) dplyr, tidyr, lubridate, and readxl packages included with the tidyverse suite of packages [@Wickham19] to assist with data processing.  Importantly, all source files were kept on Google Drive and were unaltered from their original format to isolate the original data from any downstream analysis.

Combining the source files on Google Drive into a synthesized format in R also required a consistent set of variables and naming conventions that were applied to each file.  For example, parameters included in each water quality file were converted to a common set of names and all units were standardized (e.g., total nitrogen as `tn` and mg/L, chlorophyll-a as `chla` and $\mu$g/L, etc.). This required developing custom processing code for data files from each partner to address each unique use case.  As a result, the final processing file included almost 3000 lines of code to create the synthesized data products.  Each partner agency was also assigned a unique identifier in the synthesized files so effort could be tracked and reported in the dashboard (e.g., Pinellas County as `pinco`, Florida Department of Environmental Protection as `fldep`, etc.).  All synthesized files were saved in a long and tidy format, where each observation was given its own row, each variable was in its own column, and only one value was assigned to a cell [@Wickham14c].  This allowed easier filtering of observations of interest (as required with the dashboard) using the standardized naming conventions applied to each source file. More importantly, the developed workflow was used iteratively, often several times each week, as new data files were uploaded to the Drive folders.  This required executing the synthesis code locally (i.e., on a personal computer) before pushing data to the dashboard and constantly revising the code to address new data formats or other issues encountered when new files were uploaded.

@@

Both the server and UI components of the dashboard were developed in a single R Markdown [@Xie20] file totaling `r nlns` lines of code, with `r srvpr` and `r uipr` of the code dedicated to the server and UI, respectively.  The server components of the dashboard included a setup component (`r setlns` lines, `r setpr` of the total) and reactive component (`r rctlns`, `r rctpr` of the total).  The setup imported all R package dependencies to run the dashboard (15 total), the synthesized input data from Google Drive (20 \.RData files), an R script with custom functions to summarize or plot the results, five log files, and various "static" objects used in other parts of the dashboard that did not depend on user inputs. The rest of the server included thirty-one reactive objects (`r rctlns` lines, `r rctpr` of the total) that received different user inputs and returned a shiny data object sent back to the user interface.  The imported log files were created during the data synthesis steps external to the dashboard and were specific to each data type (e.g., water quality, phytoplankton, etc.).  Each log file was used to display on the dashboard when a data type was last modified, to inform users when updates were made to the input datasets.

The remainder of the R Markdown file was used to create the dashboard UI (`r uilns` lines, `r uipr` of total).  The flexdashboard package rendered with R Markdown was used to place shiny content in specific CSS elements for easier content navigation, such as boxes or tabs that are scaled automatically to a user's browser.  Specifically, the UI code organized the dashboard content (Figure \@ref(fig:dashboard)) into 1) a landing page when a user first accesses the website, 2) tabs for navigating to current results for data from a specific monitoring type, and 3) tabs for viewing historical baseline data for comparison with the response-based monitoring data.  The landing page included a text overview of the dashboard, a map showing the monitoring locations, and several summary boxes at the bottom showing current effort to date (e.g., number of sites, total days sampled, etc., Figure \@ref(fig:dashboard)). Each 'current data' tab was organized specifically for each monitoring data type. Although, common elements included maps and tabular summaries where the displayed data could be chosen by date ranges or selected parameters by the user.  Finally, the baseline data tab displayed multi-decadal monitoring data for water quality and seagrasses to provide additional context and comparison with the response-based monitoring data.

@@

```{r analytics, fig.cap = 'Piney Point dashboard (a) users by day and (b) cumulative users by source.  Source refers to how users accessed the dashboard (e.g., direct URL to the dashboard, email links, etc.).  Analytics were included in the dashboard beginning on May 13th, 2021.'}
knitr::include_graphics(here('figs/analytics.jpeg'))
```

>>>>

The long-term recovery of Tampa Bay, Florida represents a collaborative, multi-stakeholder effort to quantify and limit external nitrogen loading as a key pollutant impacting ecosystem health.  Nitrogen loading to the estuary has been reduced by approximately two thirds of historical conditions primarily through upgrades to wastewater treatment systems, stormwater controls, and habitat restoration projects [@Johansson91;@Lewis98;@Greening14;@Beck19].  As a result, chlorophyll-a concentrations (a proxy for phytoplankton production) have decreased, water clarity has improved, and seagrasses recovered to a historical 1950s baseline in 2016 [@Greening14;@Sherwood17].  Despite these historical gains, ongoing management challenges for protecting and restoring the natural resources of Tampa Bay persist (e.g., continuing stressors associated with climate change and increasing watershed development).  Recent seagrass losses over the 2018 to 2022 period and seasonally recurring harmful algal blooms since the mid-2000s in portions of Tampa Bay highlight contemporary challenges affecting the bay's current ecology [@Lopez16;@Tomasko20].

@@

Water quality monitoring focused on parameters indicative of elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily on total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton production, and additional parameters expected to respond to nutrient loading (e.g., secchi depth and dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected at transect locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay. Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet cover-abundance estimates at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

@@

A second synthesis challenge was unifying the multiple data types hosted on Google Drive into a usable format for the dashboard and additional analyses (Figure \@ref(fig:synthesis), data synthesis).  Several folders on Google Drive were created that included relevant files for each type of monitoring data (water quality, phytoplankton, etc.).  Each folder typically contained dozens of files saved as Google Sheets, often uploaded as a single sampling event by an individual monitoring agency.  For example, the folder for the water quality sampling included 91 separate files.  Extensive data "wrangling" was required to combine the files into a tidy format [@Wickham14c] and a custom processing script was created by TBEP using the open source R statistical programming language [@RCT20]. An RStudio project was created with full version control that included the synthesis script, `dat_proc.R`, that imported and processed all files from Google Drive into individual binary \.RData files (individual file for water quality, phytoplankton, etc.). These discrete files were later used to develop a subsequent results dashboard .  Several R packages were indispensable in the data synthesis effort and these included the 1) googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages for connecting and downloading files from Google Drive, and 2) dplyr, tidyr, lubridate, and readxl packages included with the tidyverse suite of packages [@Wickham19] to assist with data processing.  Importantly, all source files were kept on Google Drive and were unaltered from their original format to isolate the original data from any downstream analysis.

Combining the source files on Google Drive into a synthesized format in R also required a consistent set of variables and naming conventions that were applied to each file.  For example, parameters included in each water quality file were converted to a common set of names and all units were standardized (total nitrogen as `tn` and mg/L, chlorophyll-a as `chla` and $\mu$g/L, etc.). This required developing custom processing code for data files from each partner to address each unique use case.  As a result, the final processing file included almost 3000 lines of code to create the synthesized data products.  Each partner agency was also assigned a unique identifier in the synthesized files so effort could be tracked and reported in the dashboard (Pinellas County as `pinco`, Florida Department of Environmental Protection as `fldep`, etc.).  All synthesized files were saved in a long and tidy format, where each observation was given its own row, each variable was in its own column, and only one value was assigned to a cell [@Wickham14c].  This allowed easier filtering of observations of interest (as required with the dashboard) using the standardized naming conventions applied to each source file. More importantly, the developed workflow was used iteratively, often several times each week, as new data files were uploaded to the Drive folders.  This required executing the synthesis code locally (i.e., on a personal computer) before pushing data to the dashboard and constantly revising the code to address new data formats or other issues encountered when new files were uploaded.

@@

Both the server and UI components of the dashboard were developed in a single R Markdown [@Xie20] file totaling `r nlns` lines of code, with `r srvpr` and `r uipr` of the code dedicated to the server and UI, respectively.  The server components of the dashboard included a setup component (`r setlns` lines, `r setpr` of the total) and reactive component (`r rctlns`, `r rctpr` of the total).  The setup imported all R package dependencies to run the dashboard (15 total), the synthesized input data from Google Drive (20 \.RData files), an R script with custom functions to summarize or plot the results, five log files, and various "static" objects used in other parts of the dashboard that did not depend on user inputs. The rest of the server included thirty-one reactive objects (`r rctlns` lines, `r rctpr` of the total) that received different user inputs and returned a shiny data object sent back to the user interface.  The imported log files were created during the data synthesis steps external to the dashboard and were specific to each data type (water quality, phytoplankton, etc.).  Each log file was used to display on the dashboard when a data type was last modified, to inform users when updates were made to the input datasets.

The remainder of the R Markdown file was used to create the dashboard UI (`r uilns` lines, `r uipr` of total).  The flexdashboard package rendered with R Markdown was used to place shiny content in specific CSS elements for easier content navigation, such as boxes or tabs that are scaled automatically to a user's browser.  Specifically, the UI code organized the dashboard content (Figure \@ref(fig:dashboard)) into 1) a landing page when a user first accesses the website, 2) tabs for navigating to current results for data from a specific monitoring type, and 3) tabs for viewing historical baseline data for comparison with the response-based monitoring data.  The landing page included a text overview of the dashboard, a map showing the monitoring locations, and several summary boxes at the bottom showing current effort to date (number of sites, total days sampled, etc., Figure \@ref(fig:dashboard)). Each 'current data' tab was organized specifically for each monitoring data type. Although, common elements included maps and tabular summaries where the displayed data could be chosen by date ranges or selected parameters by the user.  Finally, the baseline data tab displayed multi-decadal monitoring data for water quality and seagrasses to provide additional context and comparison with the response-based monitoring data.

@@

```{r analytics, fig.cap = 'Piney Point dashboard (a) users by day and (b) cumulative users by source.  Source refers to how users accessed the dashboard (direct URL to the dashboard, email links, etc.).  Analytics were included in the dashboard beginning on May 13th, 2021.'}
knitr::include_graphics(here('figs/analytics.jpeg'))
```
